---
title: "datapreper package: README"
output: html_document
date: "2024-12-2"
author: Gavin Heindorf
---
# **datapreper package**
* This package is meant to make routine data cleaning and data preparation tasks much easier and quicker. Each of the functions in this package can be used to accomplish what is typically a required, but mundane, and often time consuming, task. The functions in this task allow one to quickly create a large amount of unique labels given a common base, specifically useful for naming columns within a dataframe. One can also determine the amount of missing data by row or column within a vector or dataframe. There are also functions for recoding existing variables and creating dummy coded variables based off another variable. Finally, there is a function that allows one to easily identify unique sequences within a vector of values. All these functions are designed to excelerate the data cleaning and preparation stages, so more time can be spent collecting results and running analyses!

* This package can be downloaded from github, assuming you have the devtools package installed, by running the following line of code: install_github("GavinHeindorf/LIS4370_GHeindorf/datasetpreper")

## **Summary of Code and Functions**
### **Libraries**
* The only library required to use this package is ggplot2.
```{r}
library(ggplot2)
```

### **Label Creator**
* Purpose: 'label_creator()' allows one to quickly and personally create a sequence of unique labels by merging a common base(s) with each unique element from a vector.
* Arguments:
  + base: Base(s) that are common to all labels.
  + unique: Vector of unique elements for each unique label desired.
  + order: Order in which to use the unique elements provided, by default it applies unique elements in the order their provided.
  + uniquefront: T/F, should unique element be put at the front of the labels.
  + base_sep: String used to separate base elements, if multiple are provided.
  + base_to_unique_sep: String use to separate base element(s) from unique element.
  + copies: Integer, how many copies of the labels should be returned.
*Returns: 'label_creator()' returns a vector of labels (strings) with a length equal to the length of 'unique'. If 'copies' is greater than 1, then a list with the number of elements equal to the number of copies requested will be returned, each element being a vector according to the description provided above.
  
```{r}
label_creator <- function(base,
                          unique = NULL,
                          order = 1:length(unique),
                          uniquefront = FALSE,
                          base_sep = " ",
                          base_to_unique_sep = " ",
                          copies = 1){

  ###Argument Checking
  if (missing(base)){
    stop("Argument 'base' is missing without default")
  }
  stopifnot(is.numeric(order), is.character(base_sep),
            is.character(base_to_unique_sep), is.numeric(copies),
            is.logical(uniquefront), copies >= 1)
  if (!(is.null(unique)) & length(order) != length(unique)){
    stop("Argument 'order' must be same length as 'unique'")
  }
  if (length(base_to_unique_sep) != 1 &
      length(base_to_unique_sep) != length(unique)){
    stop("Argument 'base_to_unique_sep' does not equal 1 or length of 'unique'")
  }

  order = round(order)

  ###Functional Code
  new_labels <- c()
  if (!(is.null(unique))){
    if (uniquefront){
      for (i in order){
        new_labels <- c(new_labels, paste(unique[i],
                                  paste(base, collapse = base_sep),
                                  sep = base_to_unique_sep))
      }
    } else{
      for (i in order){
        new_labels <- c(new_labels,
                        paste(paste(base, collapse = base_sep),
                                  unique[i],
                                  sep = base_to_unique_sep))
      }
    }
  } else {
    order = 1
      for (i in order){
        new_labels <- c(new_labels, paste(base, collapse = base_sep))
      }
  }
  ###Finalize copies and return
  if (copies != 1){
    final_copies <- list()
    for (i in 1:copies){
      final_copies[[i]] <- new_labels
    }
    return(final_copies)
  } else {
    return(new_labels)
  }
}
```

### **Series Identifier**
* Purpose: 'series_identifier()' allows for the quick counting and location of an overlapping or non-overlapping series within a vector of elements.
* Arguments:
  + vector: Vector within which to identify the series.
  + series: Vector of elements matching the series that one wants to count or locate.
  + overlap: T/F, can series within 'vector' overlap and still count?
  + indices: T/F, do you want the indices of each identified series to be returned?
* Returns: 'series_identifier()' returns a list with 'total' equaling the number of series identified within 'vector'; and if 'indices' is True, then 'matchindices' is included in the list. 'matchindices' is itself a list, with each element being one of the respective matches and its' associated indices.

```{r}
series_identifier <- function(vector,
                              series,
                              overlap = FALSE,
                              indices = FALSE){
  if (missing(series)){
    stop("Error: No series provided")
  }
  if (missing(vector)){
    stop("Error: No vector provided")
  }
  stopifnot(is.vector(vector), is.logical(overlap))
  if (length(series) >= length(vector)){
    stop("Error: Series provided is longer than or the same length as the vector provided")
  }

  slength <- length(series)
  inds <- 1:length(vector)
  keepers <- 1:slength

  names(vector) <- inds

  totalmatches <- 0
  foundindices <- list()

  if (overlap){
    i = 1
    while (length(keepers) >= slength){
      curset <- vector[as.character(i:(i+slength-1))]
      if (setequal(curset, series)){
        totalmatches <- totalmatches + 1
        if (indices){foundindices[[totalmatches]] <-
          as.character(i:(i+slength-1))}
      }
      keepers <- setdiff(inds, 1:i)
      vector <- vector[as.character(keepers)]
      i <- i + 1
    }
  } else {
    i = 1
    while (length(keepers) >= slength){
      curset <- vector[as.character(i:(i+slength-1))]
      if (setequal(curset == series)){
        keepers <- setdiff(inds, c(1:(i+slength-1)))
        totalmatches <- totalmatches + 1
        if (indices){foundindices[[totalmatches]] <-
          as.character(i:(i+slength-1))}
        vector <- vector[as.character(keepers)]
        i <- i + (slength)
      } else{
        keepers <- setdiff(inds, 1:i)
        vector <- vector[as.character(keepers)]
        i <- i + 1
      }
    }
  }
  if (indices){return(list(total = totalmatches,
                           matchindices = foundindices))}
  else{return(list(total = totalmatches))}
}
```

### **Recode Values**
* Purpose: 'recode_values()' allows one to easily reverse code, or otherwise recode a variable by specifying new values for each of the original values within a vector.
* Arguments:
  + variable: Vector containing the values to be recoded.
  + reverse: T/F, should values simply be recoded? Default is True. Assumes current levels are numeric.
  + curlevels: Optional: vector containing the current levels of your variable.
  + newlevels: Optional: vector containing the new levels for the variable, should correspond directly with 'curlevels'.
  + labels: Optional: if you would like the returned variable to be a factor with labels, provide a vector containing labels equal to the length of 'newlevels'.
  + remove: Level(s) from original variable that you would like removed before the variable is recoded (do not include the level(s) in the 'curlevels' or 'newlevels' arguments.
* Returns: 'recode_values()' returns the original variable recoded and formatted as a factor, without labels if none are provided for 'labels' in the function call.

```{r}
recode_values <- function(variable,
                          reverse = TRUE,
                          curlevels = FALSE,
                          newlevels = FALSE,
                          labels = FALSE,
                          remove = FALSE){

  if (missing(variable)){
    stop("Error: No variable provided")
  }

    if (!isFALSE(remove)){
    stopifnot(is.vector(remove), remove %in% variable,
              (length(variable)-length(remove)) > 0)
    if (!isFALSE(curlevels)){
      stopifnot((length(unique(variable))-length(remove)) == length(curlevels),
                all(curlevels %in% setdiff(variable, remove)))
    }
    newvar <- variable[which(variable %in% setdiff(variable, remove))]
  } else {
    newvar <- variable
  }

  if (!isFALSE(curlevels) | !isFALSE(newlevels)){
    stopifnot(is.vector(curlevels), is.vector(newlevels),
              length(curlevels) == length(newlevels),
              length(curlevels) > 0)
    reverse <- F
  }

  stopifnot(is.logical(reverse))

  if (reverse){
    stopifnot(is.numeric(variable))
    if (!isFALSE(labels)){
      print("WARNING: Labels is being ignored, values are being reverse coded")
    }
  }

  if (!isFALSE(labels)){
    stopifnot(is.vector(labels))
    if (!isFALSE(curlevels)){
      stopifnot(length(labels) == length(newlevels))
    }
  }


  if (!isFALSE(curlevels)){
    for (i in 1:length(curlevels)){
      newvar[which(newvar == curlevels[i])] <-
        newlevels[i]
    }

    if (!isFALSE(labels)){
      return(factor(x = newvar, levels = newlevels,
                  labels = labels))
    } else{
      return(factor(x = newvar, levels = newlevels))
    }
  } else if (reverse){
    curs <- order(levels(factor(newvar)))
    curs <- rev(curs)
    indices <- list()
    final <- newvar
    for (i in 1:length(curs)){
      indices <- which(newvar == curs[i])
      final[indices] <- rev(curs)[i]
    }
    return(factor(x = final))
  }
}
```

### **Missingness**
* Purpose: 'missingness()' provides one the opportunity to very easily summarize the amount of missing values in their data and to conveniently remove variables or observations that are not full cases.
* Arguments:
  + data A dataframe or vector which is to be checked for missing data.
  + missing: A single value or special character (default = NA) that is to be treated as a missing value (e.g., 0, NULL, 99).
  + returnFulls: T/F, would you like only full instances, those with no missing data, to be returned?
  + histogram: T/F, should a histogram illustrating the amount of missing data per observation or variable be printed? (only an option if 'data' is a dataframe or matrix)
  + axis: 1 = determine missingness by row; 2 = determine missingness by column. (only necessary if 'data' is a dataframe or matrix)
* Returns: 'missingness()' returns a numerical value equal to the number of missing data points if 'data' is a vector. If 'data' is a matrix or dataframe, a vector equal to the number of rows (axis = 1) or columns (axis = 2) is returned with the number of missing values for each instance or variable. If 'returnFulls' is True, then a list is returned with 'missingness' containing the vector of counts for eac  instance or variable, and 'newdata'-consisting of all instances or variables from the original dataset that had no missing values.

```{r}
missingness <- function(data, missing = NA,
                        returnFulls = FALSE,
                        histogram = FALSE,
                        axis = 1){

  if (is.vector(data)){
    returnFulls = F
    if (histogram){
      warning("Ignorning histogram = T because 'data' is vector")
      histogram = F
    }
  } else if (is.data.frame(data) | is.matrix(data)){
    stopifnot(axis %in% c(1, 2))
  } else{
    stop("ERROR: 'data' is not a dataframe, matrix, or vector")
    }

  stopifnot(is.logical(histogram), is.logical(returnFulls))


  if (is.vector(data)){
    if (is.na(missing)){
      return(sum(is.na(data)))
    } else {
      return(length(which(data == missing)))
    }
  } else{

    if (is.na(missing)){
      res <- apply(t(apply(data, axis, is.na)), axis, sum)
    } else {
      counter <- function(x, miss){return(length(which(x == miss)))}
      res <- apply(data, axis, counter, miss = missing)
    }

    if (returnFulls){
      fulls <- which(res == 0)
      if (axis == 1){
        toreturn <- list(newdata = data[fulls,], missingness = res)
      } else {
        toreturn <- list(newdata = data[fulls], missingness = res)
      }
    } else {
      toreturn <- res
    }

    if (histogram){

      posbins <- length(unique(res))
      if (posbins > 30){
        posbins <- 30
      }

      print(ggplot2::ggplot(data.frame(xy = res), aes(x = xy))+
              geom_histogram(bins = posbins, color = "red", fill = "grey57")+
              theme_classic()+
              labs(x = "Number of Missing Values", y = "Frequency",
                  title = "Distribution of Missing Values within Dataframe"))
    }

    return(toreturn)
  }
}
```

### **Dummy Coder**
* Purpose: 'dummy_coder()' allows one to quickly and easily create a dummy coded variable based off another vector.
* Arguments:
  + regexps: A vector of strings containing the regular expressions for each respective level one would like to designate in the new dummy coded variable.
  + baseoff: The vector to which the regular expressions are applied to create the dummy coded variable.
  + levels: The levels, each of which should correspond to one of the 'regexps' provided, that will be included in the new dummy coded variable.
  + type: What variable type would you like the dummy coded variable to be returned as? Options: factor, integer, numeric, or character.
  + labels: If 'type' is factor, optional labels for the levels of the new dummy coded variable.
* Returns: 'dummy_coder()' returns the dummy coded variable as a vector, type determined by 'type', that contains each of the corresponding values based off 'baseoff'.

```{r}
dummy_coder <- function(regexps,
                        baseoff,
                        levels,
                        type = "factor",
                        labels = FALSE){

  stopifnot(!missing(regexps), length(regexps) > 0,
            is.vector(baseoff), !missing(levels), is.vector(baseoff),
            length(regexps) == length(levels))

  if (!(type %in% c("factor", "numeric", "integer", "character"))){
    stop("ERROR: 'type' must be one of 'factor', 'numeric', 'integer', or 'character'")
  } else{
    if (identical(type, "numeric")){
      labels = F
      stopifnot(is.numeric(try(as.numeric(levels),
                               silent = T)))
    } else if (identical(type, "integer")){
      stopifnot(is.integer(try(as.integer(levels),
                               silent = T)))
      labels = F
    } else if (identical(type, "character")){
      stopifnot(is.character(try(as.character(levels),
                                 silent = T)))
      labels = F
    }
  }

  if (!isFALSE(labels)){
    stopifnot(is.vector(labels), length(labels) == length(levels))
  }

  newvec <- rep(NA, length(baseoff))
  remslabs <- c()

  for (i in 1:length(regexps)){
    inds <- grep(regexps[i], baseoff)
    if (length(i) == 0){
      cat("WARNING: no matches for '", regexps[i], "' in 'baseoff' vector")
      remslabs <- c(remslabs, i)
    } else {
      newvec[inds] <- levels[i]
    }
  }

  if (identical(type, "factor")){
    if (length(remslabs) > 0){levels <- levels[-remslabs]}
    if (!isFALSE(labels)){
      if (length(remslabs) > 0){labels <- labels[-remslabs]}
      finalvec <- factor(newvec, levels = levels, labels = labels)
    } else{
      finalvec <- factor(newvec, levels = levels)
    }
  } else if (identical(type, "numeric")){
    finalvec <- as.numeric(newvec)
  } else if (identical(type, "integer")){
    finalvec <- as.integer(newvec)
  } else if (identical(type, "character")){
    finalvec <- as.character(newvec)
  }

  return(finalvec)
}
```
