---
title: "Heindorf: Final Project Functions"
output: html_document
date: "2024-11-17"
author: Gavin Heindorf
---

## **Libraries**
```{r}
library(tidyverse)
library(stringr)

```

## **Label Creator**
* Purpose: can create a large number of unique labels very fast given a base and unique component to combine
* Arguments:
  + base: the stem(s) to use for all labels
  + unique: the optional unique component for each individual label
  + order: order in which unique components are used
  + uniquefront: (T/F) should unique component be put in front
  + base_sep: if multiple base components are provided, how to separate them
  + base_to_unique_sep: how to separate base component from unique component
  + copies: how many copies of all the labels should be made (default = 1)
  
```{r}
label_creator <- function(base, unique = NULL, 
                          order = 1:length(unique), 
                          uniquefront = F,
                          base_sep = " ", 
                          base_to_unique_sep = " ", 
                          copies = 1){
  
  ###Argument Checking
  if (missing(base)){
    stop("Argument 'base' is missing without default")
  }
  stopifnot(is.integer(order), is.character(base_sep), 
            is.character(base_to_unique_sep), is.numeric(copies),
            is.logical(uniquefront), copies >= 1)
  if (!(is.null(unique)) & length(order) != length(unique)){
    stop("Argument 'order' must be same length as 'unique'")
  }
  if (length(base_to_unique_sep) != 1 & 
      length(base_to_unique_sep) != length(unique)){
    stop("Argument 'base_to_unique_sep' does not equal 1 or length of 'unique'")
  }
  
  ###Functional Code
  new_labels <- c()
  if (!(is.null(unique))){
    if (uniquefront){
      for (i in order){
        new_labels <- c(new_labels, paste(unique[i], 
                                  paste(base, collapse = base_sep), 
                                  sep = base_to_unique_sep))
      }
    } else{
      for (i in order){
        new_labels <- c(new_labels, 
                        paste(paste(base, collapse = base_sep), 
                                  unique[i],
                                  sep = base_to_unique_sep))
      }
    }
  } else {
    order = 1
      for (i in order){
        new_labels <- c(new_labels, paste(base, collapse = base_sep))
      }
  }
  ###Finalize copies and return
  if (copies != 1){
    final_copies <- list()
    for (i in 1:copies){
      final_copies[[i]] <- new_labels
    }
    return(final_copies)
  } else {
    return(new_labels)
  }
}
```

## **Series Identifier**
* Purpose: Identify the occurences of a series within a larger vector
* Arguments:
  + series: the series to identify within the vector
  + vector: the vector in which to identify the series
  + overlap: (T/F) can the identified series within the vector overlap with each other
  + indices: (T/F) should the indices of each series be returned

```{r}
series_identifier <- function(series, vector, 
                              overlap = F, indices = F){
  if (missing(series)){
    stop("Error: No series provided")
  }
  if (missing(vector)){
    stop("Error: No vector provided")
  }
  stopifnot(is.vector(vector), is.logical(overlap))
  if (length(series) >= length(vector)){
    stop("Error: Series provided is longer than or the same length as the vector provided")
  }
  
  slength <- length(series)
  inds <- 1:length(vector)
  keepers <- 1:slength
  
  names(vector) <- inds
  
  totalmatches <- 0
  foundindices <- list()
  
  if (overlap){
    i = 1
    while (length(keepers) >= slength){
      curset <- vector[as.character(i:(i+slength-1))]
      if (all(curset == series)){
        totalmatches <- totalmatches + 1
        if (indices){foundindices[[totalmatches]] <-
          as.character(i:(i+slength-1))}
      }
      keepers <- setdiff(inds, 1:i)
      vector <- vector[as.character(keepers)]
      i <- i + 1
    }
  } else {
    i = 1
    while (length(keepers) >= slength){
      curset <- vector[as.character(i:(i+slength-1))]
      if (all(curset == series)){
        keepers <- setdiff(inds, c(1:(i+slength-1)))
        totalmatches <- totalmatches + 1
        if (indices){foundindices[[totalmatches]] <-
          as.character(i:(i+slength-1))}
        vector <- vector[as.character(keepers)]
        i <- i + (slength)
      } else{
        keepers <- setdiff(inds, 1:i)
        vector <- vector[as.character(keepers)]
        i <- i + 1
      }
    }
  }
  if (indices){return(list(total = totalmatches, 
                           matchindices = foundindices))}
  else{return(list(total = totalmatches))}
}
```

## **Recode Values**
* Purpose: Recode a variable given a new set of values, or simple reverse code if desired
* Arguments:
  + variable: the variable with values to be recoded
  + reverse: (T/F) should values simply be reverse coded
  + curlevels: the current levels within the variable
  + newlevels: the levels to replace the current levels with
  + labels: labels to use for the new levels
  + remove: the current levels you would like removed before recoding

```{r}
recode_values <- function(variable, reverse = T, curlevels = F, newlevels = F, labels = F, remove = F){
  
  if (missing(variable)){
    stop("Error: No variable provided")
  }
  
  if (!isFALSE(curlevels) | !isFALSE(newlevels)){
    stopifnot(is.vector(curlevels), is.vector(newlevels), 
              length(curlevels) == length(newlevels), 
              length(curlevels) > 0)
    reverse <- F
  }
  
  stopifnot(is.logical(reverse))
  
  if (reverse){
    stopifnot(is.numeric(variable))
    if (!isFALSE(labels)){
      print("WARNING: Labels is being ignored, values are being reverse coded")
    }
  }
  
  if (!isFALSE(labels)){
    stopifnot(is.vector(labels))
    if (!isFALSE(curlevels)){
      stopifnot(length(labels) == length(newlevels))
    }
  }
  
  if (!isFALSE(remove)){
    stopifnot(is.vector(remove), remove %in% variable, 
              (length(variable)-length(remove)) > 0)
    if (!isFALSE(curlevels)){
      stopifnot((length(levels(factor(variable)))-length(remove)) == length(curlevels),
                all(curlevels %in% setdiff(variable, remove)))
    }
    newvar <- variable[which(variable %in% setdiff(variable, remove))]
  } else {
    newvar <- variable
  }
  
  
  if (!isFALSE(curlevels)){
    for (i in 1:length(curlevels)){
      newvar[which(newvar == curlevels[i])] <- 
        newlevels[i]
    }

    if (!isFALSE(labels)){
      return(factor(x = newvar, levels = newlevels, 
                  labels = labels))
    } else{
      return(factor(x = newvar, levels = newlevels))
    }
  } else if (reverse){
    curs <- order(levels(factor(newvar)))
    curs <- rev(curs)
    indices <- list()
    final <- newvar
    for (i in 1:length(curs)){
      indices <- which(newvar == curs[i])
      final[indices] <- rev(curs)[i]
    }
    return(factor(x = final))
  }
}
```

## **Missingness Summarizer**
* Purpose: summarize the amount of missing data within a vector or dataframe
* Arguments:
  + data: a vector or dataframe to summarize the missingness within
  + missing: the value that represents missing data (*NA* by default)
  + returnFulls: (T/F) should only full cases be returned
  + histogram: (T/F) if a dataframe, should a histogram displaying the distribution of missing data be printed
  + axis: 1 = by columns; 2 = by rows

```{r}
missingness <- function(data, missing = NA, returnFulls = F, histogram = F, 
                        axis = 1){
  
  if (is.vector(data)){
    returnFulls = F
    if (histogram){
      warning("Ignorning histogram = T because 'data' is vector")
      histogram = F
    }
  } else if (is.dataframe(data)){
    stopifnot(axis %in% c(1, 2))
  } else{
    stop("ERROR: 'data' is not a dataframe or vector")
    }
  
  stopifnot(is.logical(histogram), is.logical(returnFulls))
  
  
  if (is.vector(data)){
    if (is.na(missing)){
      return(sum(is.na(data)))
    } else {
      return(length(which(data == missing)))
    }
  } else{
    
    if (is.na(missing)){
      res <- apply(apply(data, axis, is.na), axis, sum)
    } else {
      counter <- function(x, miss){return(lenght(which(x == miss)))}
      res <- apply(data, axis, counter, miss = missing)
    }
    
    if (returnFulls){
      fulls <- which(res == 0)
      toreturn <- list(newdata = data[fulls], missingness = res)
    } else {
      toreturn <- res
    }
    
    if (histogram){
      print(ggplot(data.frame(xy = res), aes(x = xy))+geom_histogram()+theme_classic()+
              labs(x = "Number of Missing Values", y = "Frequency", 
                  title = "Distribution of Missing Values within Dataframe"))
    }
    
    return(toreturn)
  }
}
```

## **Dummy Coder**
* Purpose: create a dummy code variable using a base variable and regular expressions to characterize the levels
* Arguments:
  + regexps: regular expressions to identify each of the new levels from the base variable
  + baseoff: variable to base the dummy coding off of
  + levels: levels for the new variable, corresponding to each of the regular expressions
  + type: type of variable that should be returned, one of "factor", "numeric", "integer", or "character"
  + labels: if type to return is factor, labels to use for each level, if desired

```{r}
dummycoder <- function(regexps, baseoff, levels, type = "factor", labels = F){
  
  stopifnot(!missing(regexps), length(regexps) > 0, 
            is.vector(baseoff), !missing(levels), is.vector(baseoff),
            length(regexps) == length(levels))
  
  if (!(type %in% c("factor", "numeric", "integer", "character"))){
    stop("ERROR: 'type' must be one of 'factor', 'numeric', 'integer', or 'character'")
  } else{
    if (str_match(type, "numeric")){
      labels = F
      stopifnot(is.numeric(try(as.numeric(levels),
                               silent = T)))
    } else if (str_match(type =, "integer")){
      stopifnot(is.integer(try(as.integer(levels),
                               silent = T)))
      labels = F
    } else if (str_match(type, "character")){
      stopifnot(is.character(try(as.character(levels), 
                                 silent = T)))
      labels = F
    }
  }
  
  if (!isFALSE(labels)){
    stopifnot(is.vector(labels), length(labels) == length(levels))
  }
  
  newvec <- rep(NA, length(baseoff))
  remslabs <- c()
  
  for (i in 1:length(regexps)){
    inds <- grep(regexps[i], baseoff)
    if (length(i) == 0){
      cat("WARNING: no matches for '", regpexps[i], "' in 'baseoff' vector")
      remslabs <- c(remslabs, i)
    } else {
      newvec[inds] <- levels[i]
    }
  }
  
  if (str_match(type, "factor")){
    if (length(remslabs) > 0){levels <- levels[-remslabs]}
    if (!isFALSE(labels)){
      if (length(remslabs) > 0){labels <- labels[-remslabs]}
      finalvec <- factor(newvec, levels = levels, labels = labels)
    } else{
      finalvec <- factor(newvec, levels = levels)
    }
  } else if (str_match(type, "numeric")){
    finalvec <- as.numeric(newvec)
  } else if (str_match(type, "integer")){
    finalvec <- as.integer(newvec)
  } else if (str_match(type, "character")){
    finalvec <- as.character(newvec)
  }
  
  return(finalvec)
}
                       
```
